Sequential algorithms to split and merge terabyte-scale 3D images

### 2017-06-10 paper.tex
Better description of cluster reads

The main problem encountered in the context of this paper is the fact
that, if not ordered properly, data reads and writes might result in
extensive seek times that, as we illustrate below, drastically limit
the I/O performance. This problem is obviously related to the ordering
of data bytes in image files, which we assume is arbitrary but known
to the algorithm. The problem is also related to the geometry of the
chunks to extract or merge, which we again assume arbitrary but known
to the algorithm. In a nutshell, we are seeking for algorithms to
reduce seek times while allowing for arbitrary chunk geometries and
arbitrary byte orders in data files. The main idea of the variations
presented in the remainder of the paper is to convert data orderings
in memory before writing them on disk.

@@

A disk is modeled by its read and write rates ($\alpha_r$ and
$\alpha_w$, in B/s), its access time $\beta$ (in s) and its seek time,
a function $\sigma$ of the seek distance (in B). More precisely,
$\sigma$ would typically be in the order of $\beta$ when an opened
file is seeked to a random position, but it would be much smaller when
a file is seeked to a nearby position.

The disk model encompasses file systems, kernel modules and disk
controllers required to access the disk from a program run in user
space. In modern systems, this often includes caching at various
levels \todo{\cite{x}}, which we will model by zeroing $\alpha_r$ or
$\alpha_w$ values depending on the application.

@@

See Algorithm~\ref{algo:cluster-reads}. As many blocks as possible are
loaded in memory, concatenated in a buffer and written as sequentially
as possible from the buffer. Although the Algorithm looks very similar
to Algorithm~\ref{algo:buffered-slices}, it should be noted that in
Algorithm~\ref{algo:cluster-reads} the buffer should actually be a
more complex data structure capable of storing multiple disjoint
sequences of contiguous bytes without having to allocate memory to
store the ``holes'' between such sequences \todo{a figure showing the
  buffer and how it fillus up would help}.  \note{More on
  defragmentation might be needed here.}
\begin{algorithm}[h]
  \caption{Buffered merging from blocks using cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

>>>>

The main problem encountered in the context of this paper is the fact
that, if not ordered properly, data reads and writes might result in
extensive seek times that, as we illustrate below, drastically limit
the I/O performance. This problem is obviously related to the ordering
of data bytes in image files, which we assume is arbitrary but known
to the algorithm. The problem is also related to the geometry of the
chunks to extract or merge, which we again assume arbitrary but known
to the algorithm. In a nutshell, we are looking for algorithms to
reduce seek times while allowing for arbitrary chunk geometries and
arbitrary byte orders in data files. The main idea of the variations
presented in the remainder of the paper is to convert data orderings
in memory before writing them on disk.

@@

We use a simplistic disk model where a disk is characterized by its
read and write rates ($\alpha_r$ and $\alpha_w$, in B/s) and its
access time $\beta$ (in s). In particular, we assume that seeks
require a constant amount of time, regardless of the position seeked
to. In practice, large variations would be expected, but modeling such
variations would inevitably lead to models specific to the hardware,
file system or operating system, which we intentionally avoid
here. Thus, our goal is to find algorithms that minimize the
\emph{number} of seek operations. Likewise, in modern systems, read
and write times are greatly impacted by caching occurring at several
levels, which we do not model here.

%% and its seek time,
%% a function $\sigma$ of the seek distance (in B). More precisely,
%% $\sigma$ would typically be in the order of $\beta$ when an opened
%% file is seeked to a random position, but it would be much smaller when
%% a file is seeked to a nearby position.
%% The disk model encompasses file systems, kernel modules and disk
%% controllers required to access the disk from a program run in user
%% space. In modern systems, this often includes caching at various
%% levels \todo{\cite{x}}, which we will model by zeroing $\alpha_r$ or
%% $\alpha_w$ values depending on the application.

@@

As detailed in Algorithm~\ref{algo:cluster-reads}, cluster reads load
multiple blocks in memory, concatenate them in a buffer and write the
buffer as sequentially as possible. The buffer is a slightly complex
data structure, for instance an associative array or Python
dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory to store the
``holes'' between such sequences \todo{a figure showing the buffer and
  how it fillus up would help}.  \note{More on defragmentation might
  be needed here.}
\begin{algorithm}[h]
  \caption{Buffered merging from blocks using cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}
